Smart Representations: {R}ationality and Evolution in a Richer Environment

### 2015-12-17 paper/paper.tex
paper rewrite

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modelling at a high level of
abstraction.  Abstraction entails simplifying assumptions. Evolutionary game models heavily
simplify reality in many ways as well. Two simplifications relevant here are that the
environment is represented as a fixed, closed and immutable \emph{stage game} and that the
focus of evolutionary selection is behavior for this stage game alone. In contrast, some argue
for studying instead the evolutionary competition of general \emph{choice mechanisms} in a rich
and variable environment
\citep[e.g.,][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}. To show that
there is no tension here, this paper introduces a method for reconciling established notions of
evolutionary game theory with evolutionary selection of choice mechanisms in variable
environments \citep[see related work
by][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}

If agents deal with a rich and varied environment, they face many different choice
situations. A crucial part of any choice mechanism is a \emph{subjective representation} of the
choice situation, in particular a manner of forming preferences and beliefs about a possibly
uncertain world. This paper asks: which preference and belief representations are ecologically
valuable (i.e., lead to high fitness)? The evolution of preferences has been subject of recent
interest in theoretical economics \citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. The
particular contribution of this paper is to provide a case, spelled out formally, that
questions about the evolution of preferences cannot be separated from questions about the
evolution of beliefs. As a point in case, we demonstrate that if agents have imprecise
probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96}, preference representations in
terms of regret that deviate from the
true, objective fitness that evolutionary selection operates on can be selected for.

@@

An alternative view on rationality of choice is \emph{adaptationism}
\citep[e.g.,][]{Anderson1991:Is-human-cognit,ChaterOaksford2000:The-Rational-An,HagenChater2012:Decision-Making}. \todo{more
  references} Adaptationism aims to explain choice behavior by appeal to evolutionary
considerations: DMs have acquired choice mechanisms that outperform other manners of choice
making in a rich and variable environment. A choice mechanism could be a set of distinct
heuristics (the DM's \emph{adaptive toolbox}) that have little in common
\citep[e.g.,][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to relate to the philosophical debate about the nature of rational choice, we here suggest to
think of a choice mechanism as a map from choice situations to action choices that contains an
explicit level of subjective representation of a decision situation (see
Figure~\ref{fig:ChoiceMechanism}).

\begin{figure}
  \centering
  \begin{tikzpicture}[node distance = 2.5cm]

@@

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = action] (payoff) {\textbf{payoff} for $a$ in $d$};

@@

    \begin{pgfonlayer}{background}
             \node [rounded corners, dashed,
       very thick,draw=black!40,fit={($(situation.south)+(0,-10pt)$) 
                                     ($(mechanism.west)+(+5pt,+42pt)$) 
                                     % ($(action.north)+(0,+10pt)$) 
                                     ($(representation.west)+(-7pt,0)$)
                                   }] (box) {};
    \end{pgfonlayer}

@@

\end{tikzpicture}
  \caption{bla bla bla}
  \label{fig:ChoiceMechanism}
\end{figure}

@@

% The following subsection describes what these belief types are and how a single action
% selection function (maxmin expected utility) gives rise to an intersting parallel with
% classical choice rules.

>>>>

Evolutionary game theory has become an established philosophical tool for probing into
conceptual issues whose complexity requires mathematical modelling at a high level of
abstraction.  Abstraction entails simplifying assumptions. Evolutionary game models frequently
simplify reality in two ways that are relevant for present purposes. Firstly, the environment
is represented as a fixed, closed and immutable \emph{stage game}; secondly, the focus of
evolutionary selection is behavior for this stage game alone. In contrast, some argue for
studying instead the evolutionary competition of general \emph{choice mechanisms} in a rich and
variable environment
\citep[e.g.,][]{FawcettHamblin2013:Exposing-the-be,McNamara2013:Towards-a-Riche}. To show that
there is no tension here, this paper introduces a method for reconciling established notions of
evolutionary game theory with evolutionary selection of choice mechanisms in variable
environments \citep[see related work
by][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}.\todo{More references
  Harley 1981!!!, O Conor \dots}

If agents deal with a rich and variable environment, they face many different choice
situations. A choice mechanism associates decision situations with action choices. A crucial
part of any choice mechanism is a \emph{subjective representation} of the decision situation,
in particular a manner of forming preferences and beliefs about a possibly uncertain world (see
Figure~\ref{fig:ChoiceMechanism} below). This paper asks: which preference and belief
representations are ecologically valuable (i.e., lead to high fitness)? The evolution of
preferences has been subject of recent interest in theoretical economics
\citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. The particular contribution of this paper is
to provide a case, spelled out formally, that questions of preference evolution should take
variability in belief representation into account as well. As a point in case, we demonstrate
that if agents have imprecise probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96},
preference representations in terms of regret that deviate from the true, objective fitness
that evolutionary selection operates on can be selected for.

@@

An alternative view on rationality of choice is \emph{adaptationism}
\citep[e.g.,][]{Anderson1991:Is-human-cognit,ChaterOaksford2000:The-Rational-An,HagenChater2012:Decision-Making}. \todo{more
  references} Adaptationism aims to explain choice behavior by appeal to evolutionary
considerations: DMs have acquired choice mechanisms that have proved to be adpative in a rich
and variable environment. A choice mechanism could be a set of distinct heuristics (the DM's
\emph{adaptive toolbox}) that have little in common
\citep[e.g.,][]{TverskyKahnemann1981:The-Framing-of-,GigerenzerGoldstein1996:Reasoning-the-F,ScheibehenneRieskamp2013:Testing-the-Ada}. But
to relate to the philosophical debate about the nature of rational choice, we here suggest to
think of a choice mechanism as a map from choice situations to action choices that contains an
explicit level of subjective representation of a decision situation (see
Figure~\ref{fig:ChoiceMechanism}). We are most interested in the question which subjective
representation schemes, i.e., general ways of forming preferences and beliefs, are better than
others.

\begin{figure}
  \centering
  \begin{tikzpicture}[node distance = 2.25cm]

@@

    \node[rounded corners, draw, rectangle, minimum height = 0.75cm, text width=6cm, text
    centered, below of = action,node distance = 1.5cm] (payoff) {\textbf{payoff} for $a$ in $d$};

@@

    \begin{pgfonlayer}{background}
             \node [rounded corners, dashed,
       very thick,draw=black!40,fit={($(situation.south)+(0,-10pt)$) 
                                     ($(mechanism.west)+(+5pt,+50pt)$) 
                                     % ($(action.north)+(0,+10pt)$) 
                                     ($(representation.west)+(-7pt,0)$)
                                   }] (box) {};
    \end{pgfonlayer}

@@

\end{tikzpicture}
  \caption{Schematic representation of action choice. A choice mechanism maps decision
    situation $d$ to action choice $a$ by way of representing $d$ in terms of subjective
    preferences and subjective beliefs.}
  \label{fig:ChoiceMechanism}
\end{figure}

Fitness of a subjective representation scheme is measured in terms of the fitness of a
choice mechanism that contains it, all else equal. Fitness of a choice mechanism $c$ is the
expected payoff of $c$ over arbitrary decision situations $d$:
\begin{align*}
  F(c) = \int P(d) \  \utils(c,d) \text{ d} d \,,
\end{align*}
where $P(d)$ is the probability (density) of $d$ occurring, i.e., our model of the statistical
properties of a rich and variable environment, and $\Utils(c,d)$ i the payoff of the
act selected by $c$ in $d$.

The same logic also applies to situations of social decision making. A game $g$ is also a
decision situation, albeit a more complex one. A choice mechanism $c$ maps $g$ onto an action
choice in $g$. The fitness of $c$ is then:
\begin{align*}
  F(c) = \int \int P(g) \  P(c') \  \utils(c,c',g) \text{ d} g \,,
\end{align*}
where $P(g)$ is the probability (density) of game $g$, $P(c')$ is the probability of
encountering a co-player who uses choice mechanism $c'$ and $\utils(c,c',g)$ is the payoff for
a $c$-player in $g$ against the $c'$-player, given by the payoff in $g$ of the actions that $c$
and $c'$ select in $g$. For the most part, we will focus on the more complex case of social
decision making here. Results about solitary decision making follow almost directly.

\begin{itemize}
\item meta-games
\end{itemize}

@@

% The following subsection describes what these belief types are and how a single action
% selection function (maxmopin expected utility) gives rise to an intersting parallel with
% classical choice rules.
